Black-swan events in animal populations

### 2016-07-05 text/ms_pnas.tex
Create pnas-formatted ms

\section{Black-swan events in animal populations}

Sean C. Anderson\textsuperscript{1*}, Trevor A.
Branch\textsuperscript{2}, Andrew B. Cooper\textsuperscript{3}, Nicholas K.
Dulvy\textsuperscript{4}

\textsuperscript{1}Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
sandrsn@uw.edu

\textsuperscript{2}School of Aquatic and Fishery Sciences, Box 355020,
University of Washington, Seattle, WA 98195, USA, e-mail:
tbranch@uw.edu

\textsuperscript{3}School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada, e-mail:
andrew\_cooper@sfu.ca

\textsuperscript{4}Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
nick\_dulvy@sfu.ca

\textsuperscript{*}Corresponding author; present address:
School of Aquatic and Fishery Sciences, Box
355020, University of Washington, Seattle, WA 98195, USA

Short title: Black-swan events in animal populations

Keywords:
Bayesian,
die-offs,
ecological surprises,
Gompertz,
heavy tails,
leptokurtic,
mass mortality,
population dynamics,
risk,
Student-$t$ distribution

Words in abstract: 141

Words in main text: 4980

Number of references: 55

Number of figures/tables: 4/0

Statement of authorship:
S.C.A. and T.A.B. conceived the project;
S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
data and wrote the paper with input from all authors.

\newpage
%125 now is 150

\section{Abstract}

Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and tend to be driven by extrinsic perturbations
such as climate, severe winters, or disease.
Thus, ignoring black-swan events will
underestimate extinction risk.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.
% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.

% \textbf{One Sentence Summary}: Black swans---statistically improbable events
% with profound consequences---occur in animal populations and understanding
% them will improve climate risk assessments.

%\clearpage

\section{Introduction}

Major surprises happen more often than expected in
financial, social, and natural systems \citep{taleb2007, sornette2009, may2008}.
Massive unpredictable market swings are responsible for the majority of
financial gains and losses \citep{taleb2007}, fatalities from the largest wars
dwarf those from all others  \citep{newman2005}, and the frequency of the most
damaging earthquakes has exceeded past expectations \citep{sornette2009}. In
ecological systems, background rates of global extinction are punctuated by
mass extinction \citep{harnik2012}, viruses can mutate suddenly to infect new
hosts, and billions of animals can die at once in mass mortality
events \citep{fey2015}.

Indeed, such die-offs may be the most important element
affecting population persistence \citep{mangel1994} and their importance is
likely to increase given projected increases in the frequency and magnitude of
climate-related extremes \citep{ipcc2012}.
% Despite this anecdotal evidence for
% ecological black swans, systematic searches of large numbers of timeseries have
% yet to reveal black-swan events \citep{keitt1998, allen2001, halley2002}.
However, the overwhelming majority of population model fitting and risk forecasting
assumes that deviations from model predictions can be represented by a normal
distribution \citep[on a log scale; e.g.][]{brook2006a, knape2012}.
If black swans occur, though, a normal distribution would under-estimate the
probability of extreme events occurring \citep{taleb2007}.

Whereas there are many reports of black-swan events,
only a flexible comparative approach consistently applied
to a large number of time series can yield insights into the frequency, strength,
and correlates of black-swan events.
We are unaware of such a comparative analysis.
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to time series
have not accounted for an underlying population dynamics model
\citep{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
sudden changes in abundance that exceed a chosen threshold of decline
\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
But this approach
does not distinguish events from
expected dynamics.
For instance, the range of natural variability 
can differ dramatically across taxa.

@@

\section{Methods}

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution.
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

We chose the $t$ distribution for a number of reasons.
Pragmatically, the $t$ distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations is not present.
Furthermore, the $t$ distribution arises from a mixture of normal distributions
with the same mean and different variances \citep{gelman2014},
which mimics the underlying basis for black-swan events.
There are a wide range of mechanistic processes
that could give rise to extreme events in population dynamics,
such as extreme climate events
\citep{meehl2004, katz2005, ipcc2012}
or sudden changes in predation from other species.
While evaluating these various hypotheses
would be important for diagnosing the underlying causes
of extreme events and to develop predictive models,
it would be impossible to tease out each of these
underlying causes from population trend data alone.
Instead we seek a simple flexible phenomenological model
that can be applied across the widest possible
range of population time series to provide a
comparative overview of the prevalence
of black-swan events.

One alternative approach would involve fitting
the generalized extreme value (GEV) distribution,
which represents the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
the GEV requires maxima per time block
or peaks beyond a chosen threshold.
Therefore, the GEV requires longer time series then typically available
for wild animal population abundance.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

We selected abundance time series from the Global Population Dynamics
Database \citep[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (Supporting Information) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated remaining missing values (sensu \citet{brook2006a}).
Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by \citet{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
population as heavy tailed.

We fit our models with Stan 2.6.0 \citep{stan-manual2015, hoffman2014}, and R
3.2.0 \citep{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again. These thresholds are in
excess of the minimums recommended by \citet{gelman2006a} of
\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
confidence intervals. In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains had sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.

\subsection{Alternative priors}

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are concerned about when categorizing
populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet[][p.~443]{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here,
and the results are therefore not shown.

\subsection{Alternative population models} We fit alternative population models to
test if four key phenomena changed our conclusions.

\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
model in which an autoregressive parameter was fit to the process-noise
residuals:
\begin{align}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align}
In addition to the parameters in the original Gompertz model, this model
estimates an additional parameter \(\phi\), which represents the correlation of
subsequent residuals. Based on the results of previous analyses with the
GPDD \citep{connors2014} and the chosen prior in previous
analyses \citep{thorson2014a}, we placed a moderately informative prior on
\(\phi\) that assumed the greatest probability density near zero with the
reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
\mathrm{max.} = 1)\).

\subsubsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \citep{segura2013}.

\subsubsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{log} abundance.

To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
zero and twice the maximum observed abundance (similar and less informative
than in \citet{delean2013}). We set the prior on
\(r_\mathrm{max}\) as uniform between 0 and 20 as in \citet{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
the Gompertz model.

\subsection{Simulation testing}
We performed two types of simulation testing.
First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
be recovered given different true values of \(\nu\) and different sample sizes.
Second, we tested the ability of the heavy-tailed Gompertz model to obtain
unbiased parameter estimates of \(\nu\) given that a set of process deviations
was provided in which the effective \(\nu\) value was close to the true \(\nu\)
value.

We found that we could consistently and precisely recover median posterior
estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~S\ref{fig:sim-nu} lower panels).

Our simulation-testing results indicate that the Gompertz model can recapture
the true value of \(\nu\) when the process noise was chosen so that appropriate
tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
of observation error caused the model to tend to underestimate the degree of
heavy-tailedness. Fitting a model with correctly specified observation error
did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from \citet{brook2006a}.
Beta regression is useful when response data are on a continuous scale
between zero and one \citep[e.g.][]{ferrari2004}. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + \bm{X_i} \bm{\beta}),
  \: \text{for } i = 1, \dots, 609\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to match the
way they are visually represented in Fig.~S\ref{fig:correlates} and to make the
relationship approximately linear on the logit-transformed response scale. All
input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in
magnitude \citep{gelman2008c}. We excluded body length as a covariate because it
was highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \citep{gelman2006c,
gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
probability density around expected parameter values while allowing for a
higher probability density far into the tails, thereby allowing the data to
dominate the posterior more strongly if it disagrees with the prior. We fit our
models with 5,000 total iterations per chain, 2500 warm-up iterations, four
chains, and discarding every second sample. We checked for chain convergence
visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
against a similar model with the glmmADMB \textsf{R} package \citep{glmmadmb}
without uncertainty around the predictors.

\subsection{Skewed Student-$t$ forecasts} To evaluate the apparent skewness
of heavy-tailed process noise we fit Gompertz models with skewed Student-$t$
distributed process noise \citep{fernandez1998}. This distribution adds one
parameter to the Student-$t$ distribution, \(\gamma\), which controls the
skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
parameter of each model at three levels of evidence for heavy-tails:
\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
(Fig.~4).

\section{Results}

@@

The majority of our heavy-tailed estimates were robust to alternative
population models, observation error, and choice of Bayesian priors. Our
conclusions were not systematically altered when we included autocorrelation
in the residuals, modelled population growth rates with or without
density dependence, or modelled the population dynamics as Ricker-logistic
instead of Gompertz (Fig.~S\ref{fig:alt}). Similarly, introducing moderate
observation error (CV = 0.2) only slightly decreased the estimated prevalence
of black-swan events (Fig.~S\ref{fig:alt}), and the strength of the prior on
\(\nu\) had little influence on estimates of black-swan dynamics
(Fig.~S\ref{fig:alt-priors}). Finally, our simulation testing shows that, if
anything, our models underpredict the true magnitude and probability of
heavy-tailed events---especially given that the time series are relatively short,
with a median of only 26
years in our analysis (Figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).

% Our results show that weaker and stronger priors have little influence on our
% conclusions about heavy-tailed dynamics (Fig.~\ref{fig:alt-priors}). When the
% data are informative about tail behaviour (i.e. when there is strong evidence
% of low \(\nu\) values, upper-right of Fig.~\ref{fig:alt-priors}), the prior has
% little impact on the estimate of \(\nu\). When the data are less informative
% about \(\nu\) (i.e.\ when there are no or few tail events and time series are
% short or noisy), the prior can pull the estimate of \(\nu\) towards larger or
% smaller values (Fig.~\ref{fig:alt-priors}). The vast majority of the
% populations with Pr\((\nu < 10)\) in the base prior were not altered
% qualitatively by this range of prior strength.

For model fits to the population data, the probability of detecting black-swan
dynamics was positively related to time-series length and negatively related to
magnitude of process noise but not clearly related to population growth rate,
density dependence, or maximum lifespan (Fig.~3b, Fig.~S\ref{fig:correlates}).
Longer time-series length was the strongest covariate of observing black-swan
dynamics: there is a 1.6 times greater probability of detecting a black-swan
event in populations with 60 time steps than in one with 30 time steps
(Fig.~S\ref{fig:correlates}).

>>>>

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

\newcommand{\tabstats}{{}{28}}
\newcommand{\tabcausessupp}{{2}}
\newcommand{\figpriors}{{1}}
\newcommand{\figsimnu}{{2}}
\newcommand{\figsimprob}{{3}}
\newcommand{\figalt}{{4}}
\newcommand{\figaltpriors}{{5}}
\newcommand{\figcorrelates}{{6}}
\newcommand{\figskewnu}{{7}}
\newcommand{\figskewprojections}{{8}}
\mathchardef\mhyphen="2D % math hyphen

\usepackage{bm}

\input{../analysis/values} % R output

\title{Black-swan events in animal populations}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b]{Sean C. Anderson}
\author[b]{Trevor A. Branch}
\author[c]{Andrew B. Cooper}
\author[a]{Nicholas K. Dulvy}

\affil[a]{Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
sandrsn@uw.edu}
\affil[b]{School of Aquatic and Fishery Sciences, Box 355020,
University of Washington, Seattle, WA 98195, USA}
\affil[c]{School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada}

% Please give the surname of the lead author for the running footer
\leadauthor{Anderson}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Black swans---statistically improbable events
with profound consequences---happen
more often than expected in financial, social, and natural systems. Our work
demonstrates the rare but systematic presence of black-swan events in animal populations
around the world (mostly birds, mammals, insects, and fishes). These events are
predominantly downward, implying that unexpected population crashes occur more
frequently than increases. Black-swan events are not driven by life history (e.g. lifespan),
but by external causes such as extreme winters and disease. Ignoring the presence of
downward black swans may severely underestimate extinction risk in animal populations,
particularly under a changing climate where such extreme events are expected to increase
in frequency and magnitude.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{S.C.A. and T.A.B. conceived the project;
S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
data and wrote the paper with input from all authors.}
\authordeclaration{Please declare any conflict of interest here.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn\@uw.edu}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{
Bayesian $|$
die-offs $|$
ecological surprises $|$
Gompertz $|$
heavy tails $|$
leptokurtic $|$
mass mortality $|$
population dynamics $|$
risk $|$
Student-$t$ distribution
}

\begin{abstract}
Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and tend to be driven by extrinsic perturbations
such as climate, severe winters, or disease.
Thus, ignoring black-swan events will
underestimate extinction risk.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}ajor surprises happen more often than expected in
financial, social, and natural systems \cite{taleb2007, sornette2009, may2008}.
Massive unpredictable market swings are responsible for the majority of
financial gains and losses \cite{taleb2007}, fatalities from the largest wars
dwarf those from all others  \cite{newman2005}, and the frequency of the most
damaging earthquakes has exceeded past expectations \cite{sornette2009}. In
ecological systems, background rates of global extinction are punctuated by
mass extinction \cite{harnik2012}, viruses can mutate suddenly to infect new
hosts, and billions of animals can die at once in mass mortality
events \cite{fey2015}.

Indeed, such die-offs may be the most important element
affecting population persistence \cite{mangel1994} and their importance is
likely to increase given projected increases in the frequency and magnitude of
climate-related extremes \cite{ipcc2012}.
% Despite this anecdotal evidence for
% ecological black swans, systematic searches of large numbers of timeseries have
% yet to reveal black-swan events \cite{keitt1998, allen2001, halley2002}.
However, the overwhelming majority of population model fitting and risk forecasting
assumes that deviations from model predictions can be represented by a normal
distribution (on a log scale; e.g.~\cite{brook2006a, knape2012}).
If black swans occur, though, a normal distribution would under-estimate the
probability of extreme events occurring \cite{taleb2007}.

Whereas there are many reports of black-swan events,
only a flexible comparative approach consistently applied
to a large number of time series can yield insights into the frequency, strength,
and correlates of black-swan events.
We are unaware of such a comparative analysis.
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to time series
have not accounted for an underlying population dynamics model
\cite{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
sudden changes in abundance that exceed a chosen threshold of decline
\cite{young1994, gerber2001, reed2003, fey2015}.
But this approach
does not distinguish events from
expected dynamics.
For instance, the range of natural variability
can differ dramatically across taxa.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution.
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

\begin{figure*}[htb]
\centering
\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
\caption{ Illustration of population dynamic models that allow for heavy
tails. (a, b) Probability density for the Student-$t$
distribution with scale parameter of $1$ and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (c to e)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels c and d represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (f to h)
Estimates of \(\nu\) from models fit to the times series in panels c to
e. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind panels f to h
illustrates the approximate region of heavy tails.}
\label{fig:1}
\end{figure*}

\section*{Results}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.3\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (a to d)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel a are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

The majority of our heavy-tailed estimates were robust to alternative
population models, observation error, and choice of Bayesian priors. Our
conclusions were not systematically altered when we included autocorrelation
in the residuals, modelled population growth rates with or without
density dependence, or modelled the population dynamics as Ricker-logistic
instead of Gompertz (Fig.~S\figalt). Similarly, introducing moderate
observation error (CV = 0.2) only slightly decreased the estimated prevalence
of black-swan events (Fig.~S\figalt), and the strength of the prior on
\(\nu\) had little influence on estimates of black-swan dynamics
(Fig.~S\figaltpriors). Finally, our simulation testing shows that, if
anything, our models underpredict the true magnitude and probability of
heavy-tailed events---especially given that the time series are relatively short,
with a median of only 26
years in our analysis (Figs~S\figsimnu,~S\figsimprob).
